Characterization of gaps in directly bonded Si compound optics

### 2014-03-04 AO_bondedSi.tex
revise History section- replace table of papers with narative

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D,2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M,2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y,2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by reducing the thickness and mass needed for the development of a thin (0.5$-$10 mm) Silicon diffractive optic that can later be bonded to a fine polished Si prism substrate.  Other groups producing SI diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

@@

We describe a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere \cite{1992JEMat..21..669M}, including IR interference, X-ray topography, and ultrasound microscopy.  The standard infrared interference technique is to examine the bond in reflection for the presence of Newton rings.  The presence and number of such rings can yield information about both the lateral and axial extend of any gap, as long as the gap along the direction of propagation is of the order of the vacuum wavelength or larger.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  Ultrasound microscopy exploits the density change at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

@@

Serious research on semiconductor wafer bonding started around 1989, with Lehmann et al. \cite{1989JaJAP..28L2141L} bonded GaAs and InP wafers to Si.  They detected no gaps with IR imaging.  They measured the surface energy of the bond and found a range of 40$-$600 dyne/cm.

The majority of work on bonding was carried out from 1990-2000.  The key advances included studies that probed the surface energy as a function of anneal temperature, and various probes of the bond chemistry through multiply reflection infrared spectroscopy and mass spectroscopy.  A fairly complete story was then constructed which tied together all the disparate measurements.  No where in the literature did we find a technique similar to the one presented here.  See the table below for a list of papers on Si wafer bonding and a note on their contribution.

%\caption{Literature review}
\begin{spacing}{1.0} 
\begin{center}
\begin{longtable}{ |c | p{8cm} | p{8cm} |}
    \hline
    Ref & Title & Notes \\ 
        \hline
	\cite{1986JAP....60.2987S} &  \emph{Silicon-to-silicon direct bonding method} &  Foundation paper  \\
         \hline
     \cite{1988JaJAP..27L2364S} &  \emph{Bubble-Free Silicon Wafer Bonding in a Non-Cleanroom Environment} &  Sequence of IR images showing bond propagation  \\
         \hline
     \cite{1989JaJAP..28L2141L} &  \emph{Bubble-Free Wafer Bonding of GaAs and InP on Silicon in a Microcleanroom} &  No gaps detected with IR imaging; measured surface energy of the bond 40$-$600 dyne/cm.  \\
         \hline
     \cite{Mitani1990} &  \emph{Bubble formation during silicon wafer bonding: causes and remedies} &  Bubble elimination at temperature greater than 800 C. \\
         \hline
	\cite{1992JEMat..21..669M} &  \emph{Wafer bonding technology for silicon-on-lnsulator applications: A review} &  Excellent paper.  Shows infrared imaging detected bubbles and the bubble density as a function of anneal temperature.  Directly compares the three bubble detection techniques\\
         \hline
     \cite{feijoo1994} &  \emph{Silicon wafer bonding studied by infrared absorption spectroscopy} &  Works out the chemistry of the interface, finds distinct temperature phases, hydrogen bonds above   \\          
         \hline
     \cite{1994JaJAP..33....6H} &  \emph{Silicon Wafer Direct Bonding without Hydrophilic Native Oxides} &  Shows x-ray topograph technique, hydrophobic is strong bond\\
              \hline         
     \cite{1995ApPhA..61..101R} &  \emph{Multiple internal reflection spectroscopy of bonded silicon wafers} &  Chemistry of the bond interface- Evidence for Si$-$H bonds\\
              \hline
     \cite{1995ApPhL..67.3614G} &  \emph{Self-propagating room-temperature silicon wafer bonding in ultrahigh vacuum} &  IR image of wafers bonding with gaps visible  \\
              \hline     
     \cite{1995ApPhL..67..863G} &  \emph{What determines the lateral bonding speed in silicon wafer bonding?} &  Spread speed is 10 mm/s, decreases with increasing ambient pressure; thickness has no effect.  \\
              \hline
     \cite{1996ApPhL..68.2222T} &  \emph{Surface activated bonding of silicon wafers at room temperature} &  Shows that Ar etching makes stronger bonds, shows no void from IR microscopy or KOH etching  \\     
              \hline
     \cite{1996JaJAP..35.2102R} &  \emph{Characterization of Interfaces of Directly Bonded Silicon Wafers: A Comparative Study of Secondary Ion Mass Spectroscopy Multiple Internal Reflection Spectroscopy, and Transmission Electron Microscopy} &  They figure out what is the chemistry participating in Si bond formation\\               
              \hline
     \cite{1998AnRMS..28..215G} &  \emph{Semiconductor wafer bonding} &  Annual Reviews article summarizing the state of the subfield\\ 
              \hline
     \cite{JJAP.37.4197} &  \emph{Effect of Surface Roughness on Room-Temperature Wafer Bonding by Ar Beam Surface Activation} &  Appears to only explore surface roughness of less than 1 nm RMS, which seems low compared to our wafers.\\ 
              \hline
     \cite{2000JAP....88.4404H} &  \emph{Modeling the dynamics of Si wafer bonding during annealing} &  Prescription for how long to run annealing experiments\\     
              \hline
     \cite{2000RScI...71.1869G} &  \emph{Photoacoustic imaging of voids in direct wafer bonding} &  Compares photoacoustic technique to the IR technique \\     
              \hline
     \cite{2001JOptA...3...85G} &  \emph{Optical contact and van der Waals interactions: the role of the surface topography in determining the bonding strength of thick glass plates} &  What is the effect of surface roughness on optical bonding; uses 1 nm roughness.\\
              \hline
     \cite{2001JAP....89.6013L} &  \emph{Modeling of hydrophilic wafer bonding by molecular dynamics simulations} &  Examines the effect of moisture and defects in forming bonds\\ 
          \cite{2007ApOpt..46.6793H} &  \emph{Direct bonding and beyond} &  Excellent paper.  Demonstrates Si direct bonding on thick substrates.\\ 
    \hline
\end{longtable}
\end{center}
\end{spacing}

>>>>

The University of Texas diffractive optics group produces diffraction gratings on silicon using microlithographic techniques \cite{1998SPIE.3354..201J,2010SPIE.7739E.146W}.  These gratings are used either in transmission (i.e. as grisms \cite{2008SPIE.7014E..77D, 2010SPIE.7739E.123G}), or in reflection (i.e. as immersion gratings, \cite{2007ApOpt..46.3400M, 2010SPIE.7739E.146W,2012SPIE.8450E..2SG}.  These gratings operate throughout the infrared, from 1.2 to 38 $\mu$m, and find application in ground-based (IGRINS \cite{2010SPIE.7735E..54Y, 2012SPIE.8450E..2SG}), airborne (SOFIA FORCAST \cite{2008SPIE.7014E..77D}), and space (JWST NIRCAM \cite{2005SPIE.5904...21B,2010SPIE.7739E.123G}) applications.  To date, our group has produced the gratings on large (up to 100 mm diameter and 30 mm thick) monolithic pieces of monocrystalline silicon and we shape the substrates by sawing, grinding, and polishing to form the final devices \cite{2010SPIE.7739E.146W}.  The next generation of astronomy and earth science spectrographs, however, will need optical devices larger and more precise than what can be made with existing technology.  The proposed Giant Magellan Telescope \cite{2012SPIE.8444E..1HJ} Near Infrared Spectrograph \cite{2006SPIE.6269E.143J,2010SPIE.7735E..87L} (GMTNIRS) will need immersion gratings with thicknesses up to 60 mm and lengths up to 200 mm.  Substrates of the required size cannot fit into conventional processing equipment like reactive ion etchers and electron-beam lithography tools.  Large, thick Si substrates are also prohibitively expensive to experiment on.  Silicon direct bonding \cite{1986JAP....60.2987S,2012SPIE.8450E..2TV} offers advantages by reducing the thickness and mass needed for the development of a thin (0.5$-$10 mm) Silicon diffractive optic that can later be bonded to a fine polished Si prism substrate.  Other groups producing SI diffractive optics are also experimenting with this approach (see \cite{2012SPIE.8450E..2TV} and \url{www.lumella.com}).

@@

There are two main solutions for dealing with bubbles.  We will refer to the two approaches as the \emph{strength approach} and the \emph{resilience approach}.  The \emph{strength approach} is to prevent bubbles from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach demands disciplined handling protocols, and rigorous multi-stage cleaning steps, although unconventional micro-cleanroom approaches exist \cite{1989JaJAP..28L2141L}.  The \emph{resilience approach} is simply to deal with the uncomfortable likelihood that bubbles will occur, and then do something to remove them.  The most cited way to get rid of IR bubbles is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in section \ref{secHistory}.

This paper introduces a technique for directly measuring gaps at silicon-silicon interfaces.  Other techniques for measuring and mapping gaps have already been described elsewhere \cite{1992JEMat..21..669M}, including IR interference\cite{Mitani1990}, X-ray topography\cite{1994JaJAP..33....6H}, and ultrasound microscopy\cite{2000RScI...71.1869G}.  The standard infrared interference technique is to examine the bond in reflection for the presence of Newton rings.  The presence and number of such rings can yield information about both the lateral and axial extend of any gap, as long as the gap along the direction of propagation is of the order of the vacuum wavelength or larger.  X-ray topography \cite{1992JEMat..21..669M,1994JaJAP..33....6H} exploits the change in optical density along lines of sight with gaps to allow precise and high spatial resolution measurements of the spatial extent of gaps with axial dimensions as small as a few nanometers but provides only coarse estimates of the axial extent.  Ultrasound microscopy exploits the density change at the gap interface to produce reflections of sound waves and map the lateral gap dimensions but produces limited information about the axial dimension \cite{2000RScI...71.1869G}.

@@

The foundation paper on Si bonding came in 1986\cite{1986JAP....60.2987S}.  After that, there was a sequence of papers using IR imaging to look for evidence of defects in the bond.  In some of cases, the bond propagation could be directly monitored\cite{1988JaJAP..27L2364S, 1995ApPhL..67.3614G}.  The bond process could be shown to be bubble-free\cite{1988JaJAP..27L2364S, 1989JaJAP..28L2141L}.  The bond surface energy was directly measured\cite{1989JaJAP..28L2141L}.  The causes and remedies for bubble formation were discussed as early as 1990\cite{Mitani1990}.  Specifically, the IR bubble density changes as a function of anneal temprature\cite{1992JEMat..21..669M}, with bubbles disappearing at temperatures greater than 1100$^\circ$ C.  The chemistry of the bonding interface began to be understood with infrared absorption spectroscopy and multiple internal reflection spectroscopy\cite{feijoo1994}, which lent credibility to the idea of distinct temperature phases and evidence for Si$-$H bonds\cite{1995ApPhA..61..101R}.  A robust picture of the chemistry as a function of anneal temperature was ultimate fleshed out\cite{1996JaJAP..35.2102R, 1998AnRMS..28..215G}.  Other variables potentially affecting bond strength were studied, for example surface roughness\cite{JJAP.37.4197}, surface topography \cite{2001JOptA...3...85G}, surface preparation \cite{1996ApPhL..68.2222T}, annealing time \cite{2000JAP....88.4404H}, ambient pressure and substrate thickness \cite{1995ApPhL..67..863G, 2007ApOpt..46.6793H}, moisture and defects \cite{2001JAP....89.6013L}.  No where in the literature did we find a technique similar to the one presented here, probably because the technique relies on the high precision of relatively new IR capable spectroscopy that wasn't widely available until this decade.
